---
title: "Introduction to TPP2D for 2D-TPP analysis"
author: "Nils Kurzawa"
date: "2/25/2019"
package: TPP2D
output:
    BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to TPP2D for 2D-TPP analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %VignetteEncoding{UTF-8}
bibliography: bibliography.bib
csl: cell.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract
Thermal proteome profiling (TPP) [@Savitski2014, @Franken2015] is an unbiased mass spectrometry-based method to assess protein-ligand interactions. It works by employing the cellular thermal shift assay (CETSA) [@Molina2013] on a proteome-wide scale which in brief monitors the profiles of proteins in cells over a temperature gradient and tries to detect shifts induced by ligand-protein interactions in a treatment versus a control sample. 2D-TPP represents a refined version of the assay [@Becher2016] which uses a concentration gradient of the ligand of interest over a temperature gradient. This package aims to analyze data retrieved from 2D-TPP experiments by a functional analysis approach.

# General information

This package aims at providing an analysis tool for datasets obtained with the 2D-TPP assay. Please note that methods for analyzing convential TPP datasets (e.g. single dose, melting curve approach) can be found at: https://bioconductor.org/packages/release/bioc/html/TPP.html and https://git.embl.de/childs/TPP-data-analysis/blob/master/NPARC_paper/reports/NPARC_workflow.Rmd .

This vignette is not aimed to represent an in-depth introduction to thermal proteome profiling, please refer to other sources for this purpose:     
- Original TPP paper: http://science.sciencemag.org/content/346/6205/1255784    
- 2D-TPP paper: https://www.nature.com/articles/nchembio.2185   
- review article: https://proteomesci.biomedcentral.com/articles/10.1186/s12953-017-0122-4    

# Installation

The TPP2D package is in the process of being submitted to Bioconductor. So long the development version on github can be installed as follows: 

```{r eval=FALSE}
require(devtools)
devtools::install_github("nkurzaw/TPP2D")
```

# Introduction

The 2D-TPP assay is usually set up in a way that for each temperature 5 different ligand concentrations (including a vehicle condition) are used and two adjacent temperatures each are multiplexed together and in a single mass spectrometry (MS) run. Typically up to 12 temperatures are used in total that add up to 6 MS runs (Figure 1).

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(TPP2D)
```

After having loaded `dplyr` the `TPP2D` package itself we start by loading example dataset which is supplied with the package. We therefor use the `import2dDataset` function.     
For this puporse we need to supply a config table that essentially describes which experimental conditions the different TMT labels used correspond to and supplies paths to the raw data files (note: since this example dataset is included in the package it does not contain a "Path" column, this is however mandatory if the data should be read in from external raw files).
```{r}
data("config_tab")
data("raw_dat_list")

config_tab
```


We then call the import function (note: we here supply a list of data frames for the "data" argument, replacing the raw data files that would be normally specified in the above mentioned column of the config table. If this is supplied the argument "data" can simply be ignored):
```{r, warning=FALSE}
import_df <- import2dDataset(configTable = config_tab, 
                             data = raw_dat_list,
                             idVar = "protein_id",
                             intensityStr = "signal_sum_",
                             fcStr = "rel_fc_",
                             nonZeroCols = "qusm",
                             geneNameVar = "gene_name",
                             addCol = NULL,
                             qualColName = "qupm",
                             naStrs = c("NA", "n/d", "NaN"),
                             concFactor = 1e6,
                             medianNormalizeFC = TRUE,
                             filterContaminants = TRUE)

recomp_sig_df <- recomputeSignalFromRatios(import_df)
```
Please refer to the help page of the function to retrieve in-depth description of the different arguments. Essentially the function needs to know the names or prefixes of the columns in the raw data files, that contain different informations like protein id or the raw or relative signal intensities measured for the different TMT labels.    
The imported synthetic dataset consists of 17 simulated protein 2D thermal profiles (protein1-17) and 3 spiked-in true positives (tp1-3). It represents a data frame with the columns:
```{r, echo=FALSE}
knitr::kable(tibble(
  column = colnames(recomp_sig_df),
  description = c("protein identifier",
                  "number of unique quantified peptides",
                  "number of unique spectra",
                  "gene name",
                  "temperature incubated at",
                  "experiment identifier",
                  "TMT label",
                  "RefCol",
                  "treatment concentration",
                  "raw reporter ion intensity sum",
                  "raw relative fold change compared to vehicle condition at the same temperature",
                  "log10 treatment concentration",
                  "median normalized fold change",
                  "recomputed reporter ion intensity",
                  "recomputed log2 reporter ion intensity"),
  required = c("Yes",
               "No",
               "No",
               "Yes",
               "Yes",
               "No",
               "No",
               "No",
               "No",
               "No",
               "No",
               "Yes",
               "No",
               "No",
               "Yes"))
  )
```

Here the column "required" indicated which of these columns is neccessary for usage of the downstream functions.    
    
    
We then begin our actual data analysis by fitting two competing models to each protein profile. 
```{r}
competed_models <- competeModels(
  df = recomp_sig_df,
  optim_fun_h0 = TPP2D:::min_RSS_h0,
  optim_fun_h1 = TPP2D:::min_RSS_h1_slope_pEC50,
  optim_fun_h1_2 = NULL)
```

Then we create a null model using our dataset to be able to estimate the FDR for a given F statistic in the next step.

```{r}
set.seed(12, kind = "L'Ecuyer-CMRG")
null_model <- bootstrapNull(
  df = recomp_sig_df,
  optim_fun_h0 = TPP2D:::min_RSS_h0,
  optim_fun_h1 = TPP2D:::min_RSS_h1_slope_pEC50,
  optim_fun_h1_2 = NULL,
  ncores = 1, B = 2)
```

Please note that setting B = 2 is not enough to guarantee faithful FDR estimation, this has simply been set for fast demonstration purposes. We recommend to use at least B = 20 for applications in praxis.    
    
To estimate the FDR for all given F statistics and retrieve all significant hits at a set FDR $\alpha$ we use the following functions:

```{r, warning=FALSE}
fdr_tab <- computeFdr( 
  df_out = competed_models,
  df_null = null_model)

hits <- findHits(
  fdr_df = fdr_tab,
  alpha = 0.1)

hits %>% dplyr::select(clustername, nObs, F_statistic, fdr)
```

Finally we can fit and plot proteins that have come up as significant in our analysis by using:
```{r}
plot2dTppFit(recomp_sig_df, "tp1")
```
or respectively for the H1 model:
```{r}
plot2dTppFit(recomp_sig_df, "tp1", model_type = "H1", optim_fun = TPP2D:::min_RSS_h1)
```

```{r}
sessionInfo()
```


# References